---
title: "Recitation 1"
output: html_document
date: "Last Updated: 22 January 2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load in your data

When working in Google Sheets or Microsoft Excel, it's typical you're working with a .xlsx file. In R, it's easier to work with a .csv file, which doesn't require any conversions from you. The file on Brightspace is a .csv file; if you want to look at it on Sheets or Excel, you can load it there too to make life easier. But for now, let's work in R.

First, make sure you've set your working directory - good practice! Next, you're going to read your file (countries.csv) into R. There are two ways to do this. One:

```{r}
# Since we're reading in a .csv, we'll use the function read.csv() and assign it some name.

countries <- read.csv("path/to/file") # Copy the path to your file into the function. To find the path, go to the file in your File Explorer > right ckick > copy path. Just like if you were setting your work directory.
```

The second is to go into your Environment and press "Import Dataset." Press "From text (base)" and find your file. **When the prompt box appears, make sure the "Headings" button is set to yes.**

```{r}
# To make sure everything looks good, we can check some things about the data frame. 

head(countries) # We can take a look at the first few rows of a data frame using head(). You can also specify the number of rows you want to see by adding a number: head(countries, 3)

colnames(countries) # We can look at the column names using colnames()

```

# Countries Activity - Q2

Let's take a look at the data from the 10th column (Birth rate crude (per 1000 people)). You'll see some NA values - those are missing data. 

```{r}
countries[, 10]

```

Let's plot a histogram of the data to try and understand what we're working with. We can do so using the base r hist() command.
```{r}
hist(countries[, 10])

```

Wow... that looks so good... Let's adjust the bins - hopefully they'll look a little nicer.

```{r}
hist(countries[,10], breaks = 20)
hist(countries[,10], breaks = 10)
hist(countries[,10], breaks = 5)
```

Why does it still look wonky? Take a look at the values in that column and see if you can spot what looks wrong.

If you wanted to remove a value from your table, you could do so by manually going into Excel/Sheets, deleting it, then reload your file into R... or we could just delete it in R. To remove a particular cell from a table, we can use the indexing: countries[row, column]. So to delete the value for the 5th row, and 4th column (for example), we would do:
```{r}
countries[5, 4]
countries[5, 4] <- NA
countries[5, 4] # we've now replaced the value to an NA value
```

Now do the same to the value that is causing problems for the birth rate histogram, and plot the histogram again. To save a plot, you can right click on the plot then press "Save as...". (Also, it must be said that normally you work in R in your Source, so your plot will appear in the "Plots" box in the bottom right. Here, it'll appear in the Markdown file). 
```{r}
TYPE HERE
```


# Bat Activity - Q3
A slight detour. Let's look at the data from the bat_tongues.csv file. Read it in and double check the columns by using head() or columns(). In my case, I've read the file in as bat_tongues. You can adjust the names in the box below. From there, we'll plot a scatterplot. 

```{r}
x <- bat_tongues[, 2] # we'll use palate length as the independent variable
y <- bat_tongues[, 3] # we'll use tongue length as our dependent variable

plot(x, y) # You can assign the values to plot them, or you can just put them in directly below.
plot(bat_tongues[, 2], bat_tongues[, 3])
```


# Back to the Countries Dataset - Q4
Plot the relationship between:
- personal computers and life expectancy
- number of physicians and life expectancy

The actual column names are:
- "Personal_computers_per_100_people_"
- "Life_expectancy_at_birth_total_years_"
- "Physicians_per_1000_people_"
The names are funky because I had to adjust some things. Oops!

When you do so, make sure to remove any outliers so that your plots don't look too weird!

```{r}
TYPE HERE

```


